---
title: "Microbial Innovations"
author: "Mario E. Muscarella, James P. O'Dwyer"
date: "Last updated on `r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
  - \usepackage{graphics}
output: 
  pdf_document: 
    fig_caption: yes
---


# Initial Setup
```{r}
# Clear Environment and Set Working Directory
rm(list=ls())
setwd("~/GitHub/microbial-innovations/analyses")

# Add Basic Functions
se <- function(x, ...){sd(x, na.rm = TRUE)/sqrt(length(na.omit(x)))}
ci <- function(x, ...){1.96 * sd(x,na.rm = TRUE)}

# Load Packages
require("ape"); require("png"); require("grid")
require("ggtree")
require("phytools")
library("colorspace")
require("simecol")
require("picante")
require("tidyr")

# Load Source Functions
source("../bin/MothurTools.R")
source("../bin/di2multi2.R")
source("../bin/TraitMap.R")

# Define Simple Functions
ggplotColours <- function(n = 6, h = c(0, 360) + 15){
  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
}

ttest <- function(reg, coefnum, val){
  co <- coef(summary(reg))
  tstat <- (co[coefnum,1]-val)/co[coefnum,2]
  pstat <- 2 * pt(abs(tstat), reg$df.residual, lower.tail = FALSE)
  return(list = c(t = tstat, df = reg$df.residual, p =  pstat))
}

# Save Default Plot Settings
opar <- par(no.readonly = TRUE)  # Saves plot defaults

# Run All?
run.all <- FALSE
```

## Import Green Genes Tree File
```{r}
GG <- read.tree("../data/IMG.tree")
```

## Import Taxonomy Information
```{r}
GG.tax <- read.tax("../data/IMG.pds.wang.taxonomy", format = "rdp", col.tax = 2, tax.levels = 6)

```

# Edge Length Patterns
## Edge Length Distribution
```{r, results = "hide"}
GG.edges <- GG$edge.length

png(filename="../figures/GG_BranchLength.png",
    width = 800, height = 800, res = 96*2)

par(mar = c(5, 6, 3, 1) + 0.1, oma = c(1,1,1,1))

hist(log10(GG.edges), main = "Green Genes: IMG Isolates",
     axes = F, xlab = "", ylab = "", 
     xlim = c(-4, 0), ylim = c(0, 800), 
     col = "gray")


axis(1, at = c(-4, -3, -2, -1, 0, 1), 
     labels = expression(10^-4, 10^-3, 10^-2, 10^-1, 1, 10^1),  
     las = 1, lwd = 1.5)
axis(2, labels = T, las = 1, lwd = 1.5)

mtext("Edge Length", side = 1, cex = 1.25, line = 2.5)
mtext("Frequency", side = 2, cex = 1.25, line = 3.5)


# Close Plot Device
dev.off()
graphics.off()
#```
## Show Plot
#```{r}
img <- readPNG("../figures/GG_BranchLength.png")
grid.raster(img)
```

## Edge Length vs Support
```{r}
node.tab <- data.frame(matrix(NA, GG$Nnode, 5))
colnames(node.tab) <- c("node", "support", "edge_up", "edge_down1", "edge_down2")
node.tab$node <- as.numeric(c(1:GG$Nnode))
node.tab$support <- GG$node.label
node.tab[1,3] <- NA
for (i in 2:GG$Nnode){
  edge.up <- which(GG$edge[,2] == length(GG$tip.label) + i)
  node.tab[i,3] <- GG$edge.length[edge.up]
  edge.down <- which(GG$edge[,1] == length(GG$tip.label) + i)
  node.tab[i,4] <- GG$edge.length[edge.down[1]]
  node.tab[i,5] <- GG$edge.length[edge.down[2]]
}

node.tab$edge_down_min <- apply(cbind(node.tab$edge_down1, node.tab$edge_down2), 1, min)

scatterhist <- function(x, y, xlab="", ylab=""){
  zones=matrix(c(2,0,1,3), ncol=2, byrow=TRUE)
  par(oma = c(1,1,1,1))
  layout(zones, widths=c(4/5,1/5), heights=c(1/5,4/5))
  xhist = hist(x, plot=FALSE)
  yhist = hist(y, plot=FALSE)
  top = max(c(xhist$counts, yhist$counts))
  par(mar=c(5,5,1,1))
  plot(x,y, las = 1, xlab = "", ylab = "")
  mtext(xlab, side=1, line=3, outer=F, cex = 1.25)
  mtext(ylab, side=2, line=3, outer=F, cex = 1.25)
  par(mar=c(0,5,1,1))
  barplot(xhist$counts, axes=FALSE, space=0)
  par(mar=c(5,0,1,1))
  barplot(yhist$counts, axes=FALSE, space=0, horiz=TRUE)
}
```

```{r, results = "hide"}
png(filename="../figures/NodeSupport_GG.png",
    width = 800, height = 800, res = 96*2)

# Upstream
scatterhist(x = log10(node.tab$edge_up), y = as.numeric(node.tab$support),
            xlab = "Upstream Edge Length", ylab = "Node Support")

# Downstream
scatterhist(log10(node.tab$edge_down_min), as.numeric(node.tab$support),
            xlab = "Downstream Edge Length", ylab = "Node Support")

mean(as.numeric(node.tab$support), na.rm = T)
mean(as.numeric(node.tab$support[log10(node.tab$edge_up) < -1.75]), na.rm = T)
mean(as.numeric(node.tab$support[log10(node.tab$edge_up) > 0.75]), na.rm = T)
mean(as.numeric(node.tab$support[log10(node.tab$edge_up) < -2.25]), na.rm = T)

# Close Plot Device
dev.off()
graphics.off()
#```
## Show Plot
#```{r}
img <- readPNG("../figures/NodeSupport_GG.png")
grid.raster(img)
```

# GreenGenes Tree
## GG Exploratory Tree with Domain Grouping
```{r, results = "hide"}
png(filename="../figures/GG_explore.png",
    width = 1600, height = 1600, res = 96*2)

GG.2 <- GG
GG.tax2 <- GG.tax[match(GG.2$tip.label, GG.tax$OTU), ]
groupInfo <- split(GG.2$tip.label, GG.tax2$Domain)
GreenTree <- groupOTU(GG.2, groupInfo, group_name = "Domain")
levels(attributes(GreenTree)$Domain) <- names(groupInfo)
ggtree(GreenTree, aes(color = Domain), layout="rectangular") + 
  geom_text(aes(label = node), show.legend = FALSE,
            hjust=-0.1, vjust = -0.5, size = 2) + 
  theme(legend.position="top", legend.key = element_rect(colour = NA)) + 
  geom_point2(aes(subset=(node == 3518)), size=3, shape=18, 
                 colour=ggplotColours(n = 4)[1], fill = "gray80") + 
  scale_color_manual(values=c(ggplotColours(n = 4)[2:3])) 

# Close Plot Device
dev.off()
graphics.off()

# Clean Up
#```
## Show Plot
#```{r}
img <- readPNG("../figures/GG_explore.png")
grid.raster(img)
```


## Reroot GreenGenes Tree
```{r, results = "hide"}
png(filename="../figures/GG_root.png",
    width = 1600, height = 1600, res = 96*2)

GreenTree.r <- midpoint.root(GreenTree)
groupInfo <- split(GG.2$tip.label, GG.tax2$Domain)
GreenTree.r2 <- groupOTU(GreenTree.r, groupInfo, group_name = "Domain")
levels(attributes(GreenTree.r2)$Domain) <- c("root", names(groupInfo))
rt <- ggtree(GreenTree.r2, aes(color = Domain), layout="rectangular")
rt + geom_point2(aes(subset=(node == 3518)), size=4, shape=18, 
                 colour=ggplotColours(n = 4)[1], fill = "gray80") + 
  theme(legend.position="top", legend.key = element_rect(colour = NA)) + 
  scale_color_manual(values=c(ggplotColours(n = 4)[1:3])) 

# Close Plot Device
dev.off()
graphics.off()

# Clean Up
rm(rt)
#```
## Show Plot
#```{r}
img <- readPNG("../figures/LTP_root.png")
grid.raster(img)
```

## Export Rooted Tree
```{r}
write.tree(GreenTree.r, file = "../data/GG.rooted.tre")
write.nexus(GreenTree.r, file = "../data/GG.rooted.nex")
```

## GreenGenes Exploratory Tree with Cyano Grouping
```{r, results = "hide"}
png(filename="../figures/GG_Cyano.png",
    width = 1600, height = 1600, res = 96*2)

GG.tax <- read.tax("../data/IMG.pds.wang.taxonomy", col.tax = 2, tax.levels = 6)
GG.tax.rooted <- GG.tax[match(GreenTree.r$tip.label, GG.tax$OTU), ]
all.equal(as.factor(GG.tax.rooted$OTU), as.factor(GreenTree.r$tip.label))

GG.cyano <- GG.tax.rooted$Phylum

for (i in 1:length(GG.cyano)){
  if (GG.cyano[i] == "Cyanobacteria/Chloroplast"){
    GG.cyano[i] = "Cyanobacteria"
  } else {
    if (GG.tax.rooted$Domain[i] == "Archaea"){
      GG.cyano[i] = "Archaea"
    } else {
      GG.cyano[i] = "Other Bacteria"
    }
  }
}

groupInfo <- split(GreenTree.r$tip.label, GG.cyano)

# How Many Cyanos
length(grep("[C|c]yano*", GG.tax.rooted$Phylum))

GG.cy <- groupOTU(GreenTree.r, groupInfo, group_name = "Cyano")
levels(attributes(GG.cy)$Cyano) <- c("Root", names(groupInfo))


GG <- ggtree(GG.cy, aes(color = Cyano, layout="rectangular"))

GG + geom_text(aes(label = node), show.legend = FALSE,
            hjust=-0.1, vjust = -0.5, size = 2) + 
  geom_point2(aes(subset=(node == 4587)), size=4, shape=18, 
                 colour=ggplotColours(n = 4)[1], fill = "gray80") + 
  geom_point2(aes(subset=(node == 2732)), size=4, shape=18, 
                 colour=ggplotColours(n = 4)[1], fill = "gray80") + 
  theme(legend.position="top", legend.key = element_rect(colour = NA)) + 
  scale_color_manual(values=c(ggplotColours(n = 4)[c(1,2,4,3)]), name="") 

# Close Plot Device
dev.off()
graphics.off()

# Clean Up
#```
## Show Plot
#```{r}
img <- readPNG("../figures/GG_Cyano.png")
grid.raster(img)
```

## ID Midpoint and Cyano tips
```{r}
# Root
D <- cophenetic(GreenTree.r)
dd <- max(D)
ii <- which(D == dd)[1]
ii <- c(ceiling(ii/nrow(D)), ii%%nrow(D))
if (ii[2] == 0) 
    ii[2] <- nrow(D)
spp <- rownames(D)[ii]

# Cyano
D <- cophenetic(extract.clade(GreenTree.r, node = 4587))
dd <- max(D)
ii <- which(D == dd)[1]
ii <- c(ceiling(ii/nrow(D)), ii%%nrow(D))
if (ii[2] == 0) 
    ii[2] <- nrow(D)
spp <- rownames(D)[ii]
```

## Make Tree Ultrametric
I used treePL to make the tree ultrametric. My config file was pretty simple. 
It took way too long to get treePL to work. Maybe an alternative would be better.
It looks like there are some program in R. 
The most common are chronos, chronopl, chronoMPL. But I don't know how they compare.

The output file was `LTP.dated.tree`
```{r, results = "hide"}
GG.um <- read.tree("../data/GG.dated.tree")

png(filename="../figures/GG_dated.png",
    width = 1600, height = 1600, res = 96*2)

GreenTree.um <- groupOTU(GG.um, groupInfo, group_name = "Domain")
levels(attributes(GreenTree.um)$Domain) <- c("root", names(groupInfo))
um <- ggtree(GreenTree.um, aes(color = Domain), layout="rectangular")
um + geom_point2(aes(subset=(node == 4587)), size=4, shape=18, 
                 colour=ggplotColours(n = 4)[4], fill = "gray80") + 
  geom_point2(aes(subset=(node == 2732)), size=4, shape=18, 
                 colour=ggplotColours(n = 4)[1], fill = "gray80") + 
  theme(legend.position="top", legend.key = element_rect(colour = NA)) + 
  scale_color_manual(values=c(ggplotColours(n = 4)[c(1,2,4,3)]), name="") 

# Close Plot Device
dev.off()
graphics.off()

# Clear Up
rm(um)
#```
## Show Plot
#```{r}
img <- readPNG("../figures/GG_dated.png")
grid.raster(img)
```

## Branch Lenght Comparision
```{r, results = "hide"}
png(filename="../figures/GG_Branch_Comparision.png",
    width = 800, height = 800, res = 96*2)

reg <- GreenTree.r2$edge.length
um <- GreenTree.um$edge.length

# Are nodes the same?
sum(GreenTree.r2$node.label != GreenTree.um$node.label)

zero.branch <- which(reg == 0 | um == 0)
reg.l <- log10(reg[-zero.branch]) + 2.5
um.l <- log10(um[-zero.branch]) 

mod1 <- lm(um.l ~ reg.l)
summary(mod1)
ttest(mod1, 2, 1)

par(mar = c(6,6,1,1) + 0.1, oma = c(0.5,0.5,0.5,0.5))
plot(reg.l, um.l, 
     xlim = c(-2, 2.5), ylim = c(-1, 4),
     axes=F, xlab = "", ylab = "")

# abline(mod1, from = -2.5, to = 1.5, lty = 2, lwd = 2, col = "red")
new <- data.frame(reg.l = seq(0, 3.2, 0.1))
lines(x = t(new), y = predict(mod1, new), lty = 2, lwd = 2, col = "red")
clip(min(reg.l), max(reg.l), min(um.l), max(um.l))
abline(mod1$coefficients[1], 1, lty = 3, lwd = 2, col = "blue", untf = T)

axis(side = 1, las = 1, at = c(seq(-3, 1, 1) + 2.5), 
     labels = expression(10^-3, 10^-2, 10^-1, 1, 10^1))
axis(side = 2, las = 1, at = c(seq(-5,3,2)), 
     labels = expression(10^-5, 10^-3, 10^-1, 10^1, 10^3)) 

legend("bottomright", legend = c("linear model", "isometric expectation", 
                                 "slope t-test: p = < 0.001"), 
       lty = c(2, 3, 0), col = c("red", "blue", "black"), 
       bty = "n", cex = 0.6)

mtext("Non-Ultrametic Branch Length\n(Substitution Rate)", side = 1,line = 4, cex = 1.25)
mtext("Ultrametic Branch Length\n(Time)", side = 2, line = 3, cex = 1.25)

box(lwd = 2)

# Close Plot Device
dev.off()
graphics.off()

rm(mod1)
#```
## Show Plot
#```{r}
img <- readPNG("../figures/GG_Branch_Comparision.png")
grid.raster(img)
```

## Phylogeny Topological Features
```{r}
# This takes awhile, so you can turn it off

if (run.all == TRUE){
  tree.thres <- data.frame(matrix(NA, 100, 5))
  colnames(tree.thres) <- c("threshold", "nodes", "polytomies", "avg.polytomy")
  tree.thres$threshold <- seq(0.01, 0.15, length.out = 100)
  tree.thres$threshold <- 10^c(seq(-3, 0, length.out = 100))
  
  um.thres <- data.frame(matrix(NA, 100, 5))
  colnames(um.thres) <- c("threshold", "nodes", "polytomies", "avg.polytomy")
  um.thres$threshold <- 10^c(seq(-3, 2, length.out = 100))
  
  phy1 <- GreenTree.r2
  phy2 <- GreenTree.um
  
  for (i in 1:dim(tree.thres)[1]){
    phy.grained <- di2multi2(phy1, tree.thres[i,1])
    polys <- table(table(phy.grained$edge[,1]))
    poly.size <- table(phy.grained$edge[,1])
    tree.thres[i, 2] <- phy.grained$Nnode
    tree.thres[i, 3] <- sum(polys[which(as.numeric(names(polys)) > 2)]) 
    tree.thres[i, 4] <- mean(poly.size[which(poly.size > 2)])
    tree.thres[i, 5] <- length(poly.size[which(poly.size > 3)])
  }
  
  for (i in 1:dim(um.thres)[1]){
    phy.grained <- di2multi2(phy2, um.thres[i,1])
    polys <- table(table(phy.grained$edge[,1]))
    poly.size <- table(phy.grained$edge[,1])
    um.thres[i, 2] <- phy.grained$Nnode
    um.thres[i, 3] <- sum(polys[which(as.numeric(names(polys)) > 2)]) 
    um.thres[i, 4] <- mean(poly.size[which(poly.size > 2)])
    um.thres[i, 5] <- length(poly.size[which(poly.size > 3)])
  }
}
```

```{r, results = "hide"}
if (run.all == TRUE){
  png(filename="../figures/GG_Tree_Topology.png",
      width = 1600, height = 1600, res = 96*2)
  
  layout(matrix(1:6, 3))
  par(mar = c(2,4,0.5,0.5), oma = c(3,1.5,3.5,1.5))
  plot(tree.thres$nodes ~ log10(tree.thres$threshold),
       xlab = "", ylab = "", type = "l", lwd = 2,
       xaxt = "n", ylim = c(0,2000))
  axis(side = 1, at = seq(-3,0,1), labels = c(expression(10^-3, 10^-2, 10^-1, 1)))
  mtext("# Nodes", side = 2, line = 2.5)
  mtext("Original Tree", side = 3, line = 1, outer = F)
  box(lwd = 2)
  plot(tree.thres$polytomies ~ log10(tree.thres$threshold),
       xlab = "", ylab = "", type = "l", lwd = 2, 
       xaxt = "n", ylim = c(0,300))
  axis(side = 1, at = seq(-3,0,1), labels = c(expression(10^-3, 10^-2, 10^-1, 1)))
  mtext("# Polytomies", side = 2, line = 2.5)
  box(lwd = 2)
  plot(log10(tree.thres$avg.polytomy) ~ log10(tree.thres$threshold),
       xlab = "", ylab = "", type = "l", lwd = 2, 
       xaxt = "n", ylim = c(0,4))
  axis(side = 1, at = seq(-3,0,1), labels = c(expression(10^-3, 10^-2, 10^-1, 1)))
  mtext("Threshold (substitutions)", side =1, line = 2.5, cex = 1)
  mtext("Average Polytomy Size", side = 2, line = 2.5)
  box(lwd = 2)
  
  plot(um.thres$nodes ~ log10(um.thres$threshold),
       xlab = "", ylab = "", type = "l", lwd = 2,
       xaxt = "n", ylim = c(0,3000), xlim = c(-3,2))
  axis(side = 1, at = seq(-3,2,1), 
       labels = c(expression(10^-3, 10^-2, 10^-1, 1, 10^1, 10^2)))
  mtext("# Nodes", side = 2, line = 2.5)
  mtext("Ultrametric Tree", side = 3, line = 1, outer = F)
  box(lwd = 2)
  plot(um.thres$polytomies ~ log10(um.thres$threshold),
       xlab = "", ylab = "", type = "l", lwd = 2,
       xaxt = "n", ylim = c(0,300), xlim = c(-3,2))
  axis(side = 1, at = seq(-3,2,1), 
       labels = c(expression(10^-3, 10^-2, 10^-1, 1, 10^1, 10^2)))
  mtext("# Polytomies", side = 2, line = 2.5)
  box(lwd = 2)
  plot(um.thres$avg.polytomy ~ log10(um.thres$threshold),
       xlab = "", ylab = "", type = "l", lwd = 2,
       xaxt = "n", ylim = c(0,15), xlim = c(-3,2))
  axis(side = 1, at = seq(-3,2,1), 
       labels = c(expression(10^-3, 10^-2, 10^-1, 1, 10^1, 10^2)))
  mtext("Threshold (Ma)", side =1, line = 2.5, cex = 1)
  mtext("Average Polytomy", side = 2, line = 2.5)
  box(lwd = 2)
  
  
  # Close Plot Device
  dev.off()
  graphics.off()
  
  # Clean Up
}
#```
## Show Plot
#```{r}
img <- readPNG("../figures/GG_Tree_Topology.png")
grid.raster(img)
```

## Calculate Local Minimums from Edge Lengths and Maximum from Polytomies
```{r}
den.dis <- density(log10(GG.edges))
mins <- peaks(den.dis$x, den.dis$y, mode = "min")
maxs <- peaks(den.dis$x, den.dis$y, mode = "max")
```

## Obsever Polytomies through time
```{r}
if (run.all == TRUE){
  time.thres <- data.frame(matrix(NA, 100, 4))
  colnames(time.thres) <- c("threshold", "Nnodes", "Npolytomies", "avg.polytomy")
  time.thres$threshold <- 10^c(seq(-3, -1, length.out = 100))
  
  thres.info <- vector(mode = "list", length = 100)
  
  phy1 <- GreenTree.r2
  
  for (i in 1:dim(time.thres)[1]){
    phy.grained <- di2multi2(phy1, time.thres[i,1])
    poly.size <- table(phy.grained$edge[,1])
    polys <- table(table(phy.grained$edge[,1]))
    time.thres[i, 2] <- phy.grained$Nnode
    time.thres[i, 3] <- length(poly.size[which(poly.size > 2)])
    time.thres[i, 4] <- mean(poly.size[which(poly.size > 2)])
    aged <- node.age(phy.grained)
    node.info <- data.frame(matrix(NA, 4, length(aged$node.label) - 1))
    row.names(node.info) <- c("Nodes", "Node_Age", "Downstream_Branches",
                                               "Length_Upstream")
    node.info[1, ] <- as.numeric(aged$node.label[2:length(aged$node.label)])
    tips <- length(aged$tip.label)
    nodes <- aged$Nnode
    node.info[2, ] <- as.numeric(aged$ages[(tips + 2):(tips + nodes)])
    for (j in 2:length(aged$node.label)){
      node.info[3, j - 1] <- as.numeric(length(which(aged$edge[, 1] == (tips + j))))
      node.info[4, j - 1] <- as.numeric(aged$edge.length[which(aged$edge[ , 2] == (tips + j))])
    }
    thres.info[[i]] <- as.data.frame(t(as.matrix((node.info))))
  }
  
  # Explore Node Ages with Threshold
  # i = 80
  # test <- as.data.frame(thres.info[[i]])
  # plot(Node_Age ~ log10(Length_Upstream), 
  #      data = test[which(test$Downstream_Branches > 2), ])
  # legend("bottomleft",legend = paste("threshold =",
  #        round(log10(time.thres$threshold[i]), 5)), 
  #        bg = "gray", text.width = 0.5, box.col = "gray")
}


```

## Polytomy Movie
```{r}
if (run.all == TRUE){
  system("rm ../figures/movie/G*.png")
  
  threshold <- 10^c(seq(-3, -1, length.out = 50))
  
  for (i in threshold){
    thres = i
    outname <- paste("../figures/movie/G", round(thres, 4), ".png", sep = "")
    test.phy <- di2multi3(phy1 = GreenTree.r2, phy2 = GreenTree.um, 
                          tol = thres)
    poly.node <- test.phy$multi.new
    test.phy <- node.age(test.phy)
    test.phy <- groupOTU(test.phy, groupInfo, group_name = "Domain")
    levels(attributes(test.phy)$Domain) <- c("polytomy", names(groupInfo))
    
    test.tree <- ggtree(test.phy, aes(color = Domain), layout="rectangular") 
    test.tree + geom_point2(aes(subset=(node %in% c(poly.node))), size = 3, shape=18,
                  colour=ggplotColours(n = 4)[1], fill = "gray80") + 
      theme_tree(legend.position="top", legend.key = element_rect(colour = NA)) + 
      scale_color_manual(name = "", values=c(ggplotColours(n = 4)[c(1, 2, 4, 3)]), 
             labels = c("polytomy", "Archaea", "Cyanobacteria", "Other Bacteria")) + 
      annotate("text", hjust = 0, x = max(test.phy$ages) * 0.05, 
               y = length(test.phy$tip.label) * 0.95, 
               label = paste("Threshold =", round(thres, 4)))
    ggsave(outname, width = 6, height = 4, dpi = 250)
    print(paste("Threshold", round(i, 4), "done"))
  }
  
  rm(test.tree)
  
  system("convert -loop 10 -delay 100 '../figures/movie/*.png' '../figures/movie/GreenGenes.gif'")
  system("rm ../figures/movie/G*.png")
}
```

## Test Geological Window
```{r}
thres = 0.005

test.phy <- di2multi3(phy1 = GreenTree.r2, phy2 = GreenTree.um, 
                        tol = thres)
poly.node <- test.phy$multi.new
test.phy <- node.age(test.phy)
test.phy <- groupOTU(test.phy, groupInfo, group_name = "Domain")
levels(attributes(test.phy)$Domain) <- c("polytomy", names(groupInfo))
  
geo.test <- data.frame(xmin = 3200, xmax = 3600, ymin = -Inf, ymax = Inf)

test.tree <- ggtree(test.phy, aes(color = Domain), layout="rectangular")

test.tree + geom_point2(aes(subset=(node %in% c(poly.node))), size = 3, 
                        shape=18, colour=ggplotColours(n = 4)[1], 
                        fill = "gray80") + 
  theme_tree(legend.position="top", legend.key = element_rect(colour = NA)) + 
  scale_color_manual(name = "", values=c(ggplotColours(n = 4)[c(1,2,4,3)]), 
         labels = c("polytomy", "Archaea", "Cyanobacteria", "Other Bacteria")) + 
  annotate("text", hjust = 0, x = max(test.phy$ages) * 0.05, 
           y = length(test.phy$tip.label) * 0.95, 
           label = paste("Threshold =", round(thres, 4))) + 
  geom_rect(data = geo.test, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax),
              color="grey80", fill = "gray", alpha=0.5, inherit.aes = FALSE)

rm(test.tree)
```

# Traits
```{r}
IMG.traits <- read.delim("../data/predicted_patyways.L3.txt", skip = 1, 
                         row.names = 1)
dim(IMG.traits)
colnames(IMG.traits) <- gsub("X", "", colnames(IMG.traits))
sum(GreenTree.um$tip.label %in% colnames(IMG.traits))
sum(colnames(IMG.traits) %in% GreenTree.um$tip.label)

rowSums(IMG.traits > 1)

# Drop tips not found in trait table
missing <- subset(GreenTree.um$tip.label, 
                  !(GreenTree.um$tip.label %in% colnames(IMG.traits)))

IMG.tree <- drop.tip(GreenTree.um, missing)
IMG.trait.tab <- data.frame(colnames(IMG.traits), t(IMG.traits))

precal <- read.delim("../data/predicted_patyways.L3.txt", skip = 1, row.names = 1)
dim(precal)

rowSums(precal) == rowSums(IMG.traits)

summary(IMG.Cons$distance)

```


## ConsenTrait Predictions
```{r}
IMG.Cons <- read.delim("../data/IMG_ConsenTrait.txt")


# How "Deep" are traits conserved
png(filename="../figures/GG_TraitConservation.png",
      width = 800, height = 800, res = 96*2)
  
par(mar = c(4, 4, 1, 1))
plot(log10(sort(IMG.Cons$distance, decreasing = T)), 
     xlab = "", ylab = "", las = 1, col=rgb(0, 0, 1, 0.5), pch=16)

mtext(side = 1, text = "Index", line = 2.5, cex = 1.25)
mtext(side = 2, text = expression(paste("Date (log"[10], ")")), 
      line = 2.5, cex = 1.25)

# Close Plot Device
dev.off()
graphics.off()

#```
## Show Plot
#```{r}
img <- readPNG("../figures/GG_TraitConservation.png")
grid.raster(img)
```


```{r}
# Trait Origins
head(IMG.Cons)

trait.sum <- data.frame(matrix(NA, length(unique(IMG.Cons$trait)), 5))
colnames(trait.sum) <- c("Trait", "FirstDate", "MedianDate", "LateDate", "NumEvol")

for (i in 1:length(unique(IMG.Cons$trait))){
  trait <- unique(IMG.Cons$trait)[i]
  temp <- IMG.Cons[IMG.Cons$trait == trait, ]
  first.evol <- max(temp$distance)
  med.evol <- median(temp$distance)
  last.evol <- min(temp$distance)
  n.evol <- dim(temp)[1]
  trait.sum[i, 1] <- i
  trait.sum[i, 2] <- first.evol
  trait.sum[i, 3] <- med.evol
  trait.sum[i, 4] <- last.evol
  trait.sum[i, 5] <- n.evol
}

trait.sum <- trait.sum[trait.sum$NumEvol > 0, ]

png(filename="../figures/GG_TraitConservationDates.png",
      width = 1800, height = 600, res = 96*2)


layout(matrix(1:3, ncol = 3))
par(mar = c(4, 4, 2, 1), oma = c(1, 1, 1, 1))

plot(log10(trait.sum$FirstDate), jitter(trait.sum$NumEvol),
     xlab = "", ylab = "", las = 1, ylim = c(0, 80),
     col=rgb(0, 0, 1, 0.5), pch=16,
     main = "First Date of Conservation")

mtext(side = 1, text = expression(paste("Date (log"[10], ")")), 
      line = 3, cex = 1)
mtext(side = 2, text = "Conservation Points", 
      line = 2.5, cex = 1)

plot(log10(trait.sum$MedianDate), jitter(trait.sum$NumEvol),
     xlab = "", ylab = "", las = 1, ylim = c(0, 80),
     col=rgb(0, 0, 1, 0.5), pch=16,
     main = "Median Date of Conservation")

mtext(side = 1, text = expression(paste("Date (log"[10], ")")), 
      line = 3, cex = 1)
mtext(side = 2, text = "Conservation Points", 
      line = 2.5, cex = 1)



plot(log10(trait.sum$LateDate), jitter(trait.sum$NumEvol),
     xlab = "", ylab = "", las = 1, ylim = c(0, 80),
     col=rgb(0, 0, 1, 0.5), pch=16,
     main = "Latest Date of Conservation")

mtext(side = 1, text = expression(paste("Date (log"[10], ")")), 
      line = 3, cex = 1)
mtext(side = 2, text = "Conservation Points", 
      line = 2.5, cex = 1)

# Close Plot Device
dev.off()
graphics.off()

#```
## Show Plot
#```{r}
img <- readPNG("../figures/GG_TraitConservationDates.png")
grid.raster(img)
```

```{r}
trait.sum

dim(trait.sum)[1]
sum(trait.sum$NumEvol == 1)
sum(trait.sum$NumEvol > 10)
sum(trait.sum$NumEvol > 1 & trait.sum$NumEvol <= 10)



```

## Trait Types
```{r}

head(IMG.traits)[, 1:5]

traits <- rownames(IMG.traits)
trait.num <- rowSums(IMG.traits)

disease <- unique(c(grep("cancer", traits), grep("disease", traits),
                 grep("diabetes", traits), grep("losis", traits),
                 grep("ALS", traits), grep("cholerae", traits),
                 grep("plasmosis", traits)))

carb.traits <- c(grep("Glycolysis", traits), grep("Citrate", traits),
                 grep("Pentose", traits), grep("Fructose", traits),
                 grep("Galactose", traits), grep("Ascorbate", traits),
                 grep("sucrose", traits), grep("nucleotide sugar", traits),
                 grep("Pyruvate", traits), grep("Glyoxylate", traits),
                 grep("Propanoate", traits), grep("Butanoate", traits),
                 grep("C5-Branched", traits), grep("Inositol", traits))

transport.traits <- grep("transport", traits)

motility.traits <- c(grep("chemotaxis", traits), grep("motility", traits),
                     grep("Flagellar", traits))

photo.traits <- c(grep("Photosynthesis", traits))
traits[photo.traits]

dim(trait.sum)
dim(IMG.traits)


photo.sum <- trait.sum[trait.sum$Trait == photo.traits, ]
photo.Cons <- IMG.Cons[sum(IMG.Cons$trait %in% photo.traits), ]

unique(IMG.Cons$trait)[photo.traits[1]]

IMG.Cons[IMG.Cons$trait %in% unique(IMG.Cons$trait)[transport.traits], ]

```

## ASR Results
```{r}
IMG.ASR <- read.delim("../data/IMG_ASR.txt")

IMG.ASR <- IMG.ASR[complete.cases(IMG.ASR$First.Evol), ]

plot(log10(IMG.ASR$First.Evol), jitter(IMG.ASR$N.evol),
     xlab = "", ylab = "", las = 1, 
     col=rgb(0, 0, 1, 0.5), pch=16,
     main = "First Date of Conservation")

mtext(side = 1, text = expression(paste("Date (log"[10], ")")), 
      line = 3, cex = 1)
mtext(side = 2, text = "Conservation Points", 
      line = 2.5, cex = 1)

plot(log10(IMG.ASR$Med.Evol), jitter(IMG.ASR$NumEvol),
     xlab = "", ylab = "", las = 1, ylim = c(0, 80),
     col=rgb(0, 0, 1, 0.5), pch=16,
     main = "Median Date of Conservation")

mtext(side = 1, text = expression(paste("Date (log"[10], ")")), 
      line = 3, cex = 1)
mtext(side = 2, text = "Conservation Points", 
      line = 2.5, cex = 1)

```



