---
title: "Microbial Innovations: Simulations"
author: "Mario E Muscarella"
date: "Last updated on `r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
  - \usepackage{graphics}
output: 
  pdf_document: 
    fig_caption: yes
---

```{r setup, include=FALSE}
# Knitr Options
knitr::opts_chunk$set(echo = TRUE)

# Clear Environment and Set Working Directory
rm(list=ls())
setwd("~/GitHub/microbial-innovations/analyses")

# Load Packages
library("phytools")
library("ape")
library("mnormt")
library("MASS")
library("expm")
library("diagram")
library("adephylo")
require("grid")
require("png")
library("geiger")
```


# Mock Data: 
To test trait innovation approaches, we are going to work with some mock data. 
The goal is to have managable trees with traits that we can test various methods.

# Simulate Phylogenetic Tree of OTUs
```{r}
# Set Seed for Reproducibility
set.seed(199)

# Generate a tree
#tree <- pbtree(n = 100, scale = 1)
tree <- rcoal(n = 100, rooted = T)
tree$tip.label <- paste("OTU", sprintf("%05d", seq(1:100)), sep = "")

# Re Order Tree
tree <- reorder(tree)
```

# Visulaize Tree
```{r}
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 1, cex = 0.5)
```

# Define Markov Chain Trait Model Function
```{r}
MCtrait <- function(t = 5, a = 0.7, b = 0.5){
  Q <- matrix(c(a, 1-a, 1-b, b), 2, 2, byrow = T)

  x <- ((a - 1)/(a + b - 2)) * log(a + b - 1)
  y <- ((b - 1)/(a + b - 2)) * log(a + b - 1)

  H <- matrix(c(x, -x, -y, y), 2, 2, byrow = T)
  C <- eigen(H)$vectors
  D <- diag(eigen(H)$values)

  M <- expm(H * t)
  M2 <- C %*% expm(D * t) %*% ginv(C)
  
  return(list(M = M2, Q = Q))
}
```

# Generate Traits Under Markov Chain Trait Model
```{r}
# Set Seed for Reproducibility
set.seed(171)

# Generate Traits Matrix
traits <- matrix(NA, nrow = (tree$Nnode + length(tree$tip.label)), ncol = 3)
colnames(traits) <- c("Parent", "Offspring", "Trait")
traits <- as.data.frame(traits)

# Define Trait States
traitNames <- c("Off","On")

# Define Root Ancestor Traits
traits[1,] <- c("None", "101", "Off")

# Run Trait Model Given the Tree
for (i in 1:dim(tree$edge)[1]){
  traits[i +1,1:2] <- tree$edge[i, ]
  t <- tree$edge.length[i]
  p <- tree$edge[i,1]
  o <- tree$edge[i,2]
  init <- traits$Trait[which(traits$Offspring == as.character(p))]
  if (init == "Off"){u <- c(1,0)} else {
    if (init == "On"){u <- c(0,1)}}
  M <- MCtrait(t = t , a = 0.5, b = 0.55)
  prob <- round(u %*% M$M, 4)
  s <- sample(traitNames, size = 1, prob = prob)
  traits[i + 1, 1] <- p 
  traits[i + 1, 2] <- o
  traits[i + 1, 3] <- s
}

node.traits <- traits[which(as.numeric(traits$Offspring) > length(tree$tip.label)), ]
node.traits2 <- node.traits[order(as.numeric(node.traits$Offspring)), ]

tip.traits <- traits[which(as.numeric(traits$Offspring) <= length(tree$tip.label)), ]
tip.traits2 <- tip.traits[order(as.numeric(tip.traits$Offspring)), ]

n.col <- node.traits2$Trait
n.col <- gsub("On", "red", gsub("Off", "gray", n.col))

t.col <- tip.traits2$Trait
t.col <- gsub("On", "red", gsub("Off", "gray", t.col))
```

# Plot MC Model Traits on Tree
```{r}
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 2, cex = 0.3)
nodelabels(node.traits2$Offspring, cex = 0.25, frame = "circle", bg = n.col)
tiplabels(pch = 22, bg = t.col, adj = c(1.5, 0.5))
```

# Conceptual Figure for MC Trait Model
```{r}
plotmat(M$Q,pos = c(2),
        lwd = 1, box.lwd = 2,
        cex.txt = 0.8,
        box.size = 0.1,
        box.type = "circle",
        box.prop = 0.5,
        box.col = "light yellow",
        arr.type = "triangle",
        arr.length= 0.3,
        arr.width= 0.2,
        self.cex = 0.4,
        self.shifty = 0.03,
        self.shiftx = c(-0.13, 0.13),
        shadow.size = 0.001,
        main = "",
        name = c("Off", "On"))
```


# Predict Traits with Ancestral State Reconstruction
```{r}
Obs.Traits <- data.frame(OTU = tree$tip.label, Traits = tip.traits2$Trait)

# Use ACE function for Baysian Posterior Probabilities
ASR <- ace(Obs.Traits$Traits, tree, type = "d", model = "ARD", CI = T,
           marginal = T)

# Use Simmap Function to run Markov Chain Monte Carlo for Trait Statse on Edges
# MCMC For Time in each state on each edge
x <- Obs.Traits$Traits; names(x) <- tree$tip.label
ASR.2 <- make.simmap(tree, x, Q = "mcmc", model = "ARD", pi = c(1,0))
str(ASR.2)
summary(ASR.2)
plot(ASR.2)
ASR.2$maps
edge <- ASR.2$mapped.edge / tree$edge.length
# node <- rep(NA, ASR.2$Nnode)
# for (i in 2:(length(node) - 1)){
#   if (tail(names(ASR.2$maps[[i + length(ASR.2$tip.label)]]), n = 1) == "Off"){
#     node[i] <- "grey"
#   } else {
#     if (tail(names(ASR.2$maps[[i + length(ASR.2$tip.label)]]), n = 1) == "On"){
#     node[i] <- "red"
#     }
#   }
# }
```

# Plot ASR Results
```{r}
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 2, cex = 0.3)

tiplabels(pch = 22, bg = t.col, adj = c(1.5, 0.5))

nodelabels(text = node.traits2$Offspring, cex = 0.5, frame = "none", adj = c(0.5,-2.2))
nodelabels(thermo = ASR$lik.anc, cex = 0.4, piecol = c("gray", "red"))

edgelabels(pie = edge, cex = 0.4, piecol = c("gray", "red"))
```

# Define Trait Conservation Function
```{r}
# Define Function
ConsenTrait <- function(tree = "", traits = "", cutoff = 0.9){

  # Import Tree and Root  if Needed
  if (is.rooted(tree) == FALSE){
    root_tree <- midpoint.root(tree)
  } else {
    root_tree <- tree
  }
  
  # Import Traits into Function
  table <- traits
  
  # Drop tips not found in trait table
  z <- subset(tree$tip.label, !(tree$tip.label %in% table[,1]) )
  if (length(z) > 0){
    drop.tip(tree,z)
  }
  
  # Replace any negative branch lengths
  root_tree$edge.length[root_tree$edge.length <= 0] = 0.00001
  
  # ID all subtrees
  subtree <- subtrees(root_tree, wait = FALSE)

  # Initializing Results Table
  y = rep(NA, (length(subtree) * (dim(table)[2] - 1)))
  cluster_size_tab <- data.frame(trait = NA, subtree = NA, node = NA, 
                                 distance = NA, cluster_size = NA)

  # Loop Through Traits
  for (i in 2:ncol(table)){
    print(paste("Analyzing Trait", i - 1, "of", 
                ncol(table)[[1]] - 1, "...", sep = " "), quote = F)
    
    # Make Temp Table
    table_tmp <- data.frame(ID = table[, 1], Trait = table[,i])

    # Remove All Entries Not in Tree and Sort by ID
    table2 <- table_tmp[which(table_tmp$ID %in% root_tree$tip.label), ]
    table2 <- table2[sort(table2$ID), ]
    
    # Initialize Temp Result Vectors
    positives <- vector(mode = "character", length = 0)
    cluster_size <- numeric(length=0)
    cluster_dist <- numeric(length = 0)
    node_positive <- vector(mode = "character", length = 0)
    
    # Loop through all subtrees and determine if any subtrees have >90% positives
    for (j in 1:length(subtree)){
      tip_names <- subtree[[j]]$tip.label
      if (mean(table2$Trait[which(table2$ID %in% tip_names)]) > cutoff){
        match_test <- match(tip_names, positives)
        if (all(is.na(match_test))){
          positives <- c(positives,tip_names)
          node_positive <- subtree[[j]]$node.label[1]
          cluster_dist <- distRoot(subtree[[j]], tip_names, method = c("p"))
          cluster_size <- c(cluster_size, mean(cluster_dist))

          cluster_size_tab[j + length(subtree) * (i - 2), ] <- c(i - 1, j, node_positive, 
                                     mean(cluster_dist), length(cluster_dist))
          
        } else {
          if (any(is.na(match_test))) {
            print("some NAs - something is weird")
          } 
        }
      }
    }
  }
  data.out <- cluster_size_tab[complete.cases(cluster_size_tab), ]
  return(data.out)
}
```

# Run Trait Conservation Function
```{r}
# Define Traits
x.2 <- as.numeric(gsub("Off", 0, gsub("On", 1, x)))
traits <- data.frame(OTU = names(x), Traits1 = x.2, Traits2 = rev(x.2))                        
conserve <- ConsenTrait(tree, traits = traits, cutoff = 0.90)
nodes <- conserve$node[which(conserve$trait == 1)]
```

# Plot Trait Conservation Results
```{r}
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 2, cex = 0.3)
tiplabels(pch = 22, bg = t.col, adj = c(1.5, 0.5))

# nodelabels(text = node.traits2$Offspring, cex = 0.5, frame = "none", adj = c(0.5,-2.2))

nodelabels(node = c(nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(nodes), text = c(nodes), cex = 0.5, frame = "none")
```

# Generate Multipanel Figure
```{r}
png(filename="../figures/Simulation_1.png",
    width = 1600, height = 800, res = 96*2)

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(node.traits2$Offspring, cex = 0.25, frame = "circle", bg = n.col)
tiplabels(pch = 22, bg = t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- edge[(grep("*,[0-9][0-9][0-9]", rownames(edge))), ]
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = t.col, adj = c(3, 0.5))
#nodelabels(text = node.traits2$Offspring, cex = 0.5, frame = "none", adj = c(0.5,-2.2))
#nodelabels(thermo = ASR$lik.anc, cex = 0.6, piecol = c("gray", "red"))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(edge))), pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = t.col, adj = c(3, 0.5))
nodelabels(node = c(nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(nodes), text = c(nodes), cex = 0.5, frame = "none")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("../figures/Simulation_1.png")
grid.raster(img)
```


# ReRun Above Code as Function
```{r}
TraitSim <- function(a = 0.5, b = 0.55, seed1 = 199, seed2 = 171,
                     tree.method = c("coalescent", "bd")){
  # Set First Seed
  set.seed(seed1)
  
  # Generate a tree
  method <- match.arg(tree.method)
  if (method == "coalescent"){
    tree <- rcoal(n = 100, rooted = T)
    tree$tip.label <- paste("OTU", sprintf("%05d", seq(1:100)), sep = "")
  }
  if (method == "bd"){
    library("diversitree")
    tree <- rphylo(n = 100, b = 0.1, d = 0.025)
    tree$tip.label <- paste("OTU", sprintf("%05d", seq(1:100)), sep = "")
  }
  
  # Re Order Tree
  tree <- reorder(tree)
  
  # Set Second Seed
  set.seed(seed2)
  
  # Generate Traits Matrix
  traits <- matrix(NA, nrow = (tree$Nnode + length(tree$tip.label)), ncol = 3)
  colnames(traits) <- c("Parent", "Offspring", "Trait")
  traits <- as.data.frame(traits)
  
  # Define Trait States
  traitNames <- c("Off","On")
  
  # Define Root Ancestor Traits
  traits[1,] <- c("None", "101", "Off")
  
  # Run Trait Model Given the Tree
  for (i in 1:dim(tree$edge)[1]){
    traits[i +1,1:2] <- tree$edge[i, ]
    t <- tree$edge.length[i]
    p <- tree$edge[i,1]
    o <- tree$edge[i,2]
    init <- traits$Trait[which(traits$Offspring == as.character(p))]
    if (init == "Off"){u <- c(1,0)} else {
      if (init == "On"){u <- c(0,1)}}
    M <- MCtrait(t = t , a = a, b = b)
    prob <- round(u %*% M$M, 4)
    s <- sample(traitNames, size = 1, prob = prob)
    traits[i + 1, 1] <- p 
    traits[i + 1, 2] <- o
    traits[i + 1, 3] <- s
  }
  
  node.traits <- traits[which(as.numeric(traits$Offspring) > length(tree$tip.label)), ]
  node.traits2 <- node.traits[order(as.numeric(node.traits$Offspring)), ]
  
  tip.traits <- traits[which(as.numeric(traits$Offspring) <= length(tree$tip.label)), ]
  tip.traits2 <- tip.traits[order(as.numeric(tip.traits$Offspring)), ]
  
  n.col <- node.traits2$Trait
  n.col <- gsub("On", "red", gsub("Off", "gray", n.col))
  
  t.col <- tip.traits2$Trait
  t.col <- gsub("On", "red", gsub("Off", "gray", t.col))
    
  Obs.Traits <- data.frame(OTU = tree$tip.label, Traits = tip.traits2$Trait)

  # Use ACE function for Baysian Posterior Probabilities
  ASR <- ace(Obs.Traits$Traits, tree, type = "d", model = "ARD", CI = T,
             marginal = T)
  
  # Use Simmap Function to run Markov Chain Monte Carlo for Trait Statse on Edges
  # MCMC For Time in each state on each edge
  x <- Obs.Traits$Traits; names(x) <- tree$tip.label
  ASR.2 <- make.simmap(tree, x, Q = "mcmc", model = "ARD", pi = c(1,0))
  edge <- ASR.2$mapped.edge / tree$edge.length
    
  # Define Traits
  x.2 <- as.numeric(gsub("Off", 0, gsub("On", 1, x)))
  traits <- data.frame(OTU = names(x), Traits1 = x.2, Traits2 = rev(x.2))                     
  # Use ConsenTrait Function for Trait Conservation  
  conserve <- ConsenTrait(tree, traits = traits, cutoff = 0.90)
  nodes <- conserve$node[which(conserve$trait == 1)]
  
  return(list(tree = tree, node.traits2 = node.traits2, n.col = n.col,
              t.col = t.col, edge = edge, nodes = nodes))
}
```

# Run Simulation 1a (rerun of above code)
```{r}
Sim1a <- TraitSim(a = 0.5, b = 0.55, seed1 = 199, seed2 = 171,
                  tree.method = "coalescent")

png(filename="../figures/Simulation_1a.png",
    width = 1600, height = 800, res = 96*2)

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(Sim1a$node.traits2$Offspring, cex = 0.25, frame = "circle", bg = Sim1a$n.col)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- Sim1a$edge[(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), ]
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), 
           pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
nodelabels(node = c(Sim1a$nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(Sim1a$nodes), text = c(Sim1a$nodes), cex = 0.5, frame = "none")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("../figures/Simulation_1a.png")
grid.raster(img)
```

# Run Simulation 2 (new transitions, once you get the trait you keep)
```{r}
Sim1a <- TraitSim(a = 0.5, b = 0.58, seed1 = 199, seed2 = 171,
                  tree.method = "coalescent")

png(filename="../figures/Simulation_2.png",
    width = 1600, height = 800, res = 96*2)

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(Sim1a$node.traits2$Offspring, cex = 0.25, frame = "circle", bg = Sim1a$n.col)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- Sim1a$edge[(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), ]
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), 
           pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
nodelabels(node = c(Sim1a$nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(Sim1a$nodes), text = c(Sim1a$nodes), cex = 0.5, frame = "none")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("../figures/Simulation_2.png")
grid.raster(img)
```

# Run Simulation 3 (Traits really are just contemporary)
```{r}
Sim1a <- TraitSim(a = 0.5, b = 0.58, seed1 = 99, seed2 = 71,
                  tree.method = "coalescent")

png(filename="../figures/Simulation_3.png",
    width = 1600, height = 800, res = 96*2)

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(Sim1a$node.traits2$Offspring, cex = 0.25, frame = "circle", bg = Sim1a$n.col)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- Sim1a$edge[(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), ]
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), 
           pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
nodelabels(node = c(Sim1a$nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(Sim1a$nodes), text = c(Sim1a$nodes), cex = 0.5, frame = "none")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("../figures/Simulation_3.png")
grid.raster(img)
```


# Run Simulation 4 (Early Evolution with Dichotomy and Loss, ASR & C fail)
```{r}
Sim1a <- TraitSim(a = 0.58, b = 0.48, seed1 = 99, seed2 = 125,
                  tree.method = "coalescent")

png(filename="../figures/Simulation_4.png",
    width = 1600, height = 800, res = 96*2)

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(Sim1a$node.traits2$Offspring, cex = 0.25, frame = "circle", bg = Sim1a$n.col)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- Sim1a$edge[(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), ]
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), 
           pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
nodelabels(node = c(Sim1a$nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(Sim1a$nodes), text = c(Sim1a$nodes), cex = 0.5, frame = "none")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("../figures/Simulation_4.png")
grid.raster(img)
```

# Using a Birth-Death Process Tree
```{r}
Sim1a <- TraitSim(a = 0.8, b = 0.85, seed1 = 199, seed2 = 171,
                  tree.method = "bd")

png(filename="../figures/Simulation_bd1.png",
    width = 1600, height = 800, res = 96*2)

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(Sim1a$node.traits2$Offspring, cex = 0.25, frame = "circle", bg = Sim1a$n.col)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- Sim1a$edge[(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), ]
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), 
           pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
nodelabels(node = c(Sim1a$nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(Sim1a$nodes), text = c(Sim1a$nodes), cex = 0.5, frame = "none")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("../figures/Simulation_bd1.png")
grid.raster(img)
```


# Using a Birth-Death Process Tree
```{r}
Sim1a <- TraitSim(a = 0.9, b = 0.85, seed1 = 99, seed2 = 71,
                  tree.method = "bd")

png(filename="../figures/Simulation_bd2.png",
    width = 1600, height = 800, res = 96*2)

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(Sim1a$node.traits2$Offspring, cex = 0.25, frame = "circle", bg = Sim1a$n.col)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- Sim1a$edge[(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), ]
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), 
           pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
nodelabels(node = c(Sim1a$nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(Sim1a$nodes), text = c(Sim1a$nodes), cex = 0.5, frame = "none")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("../figures/Simulation_bd2.png")
grid.raster(img)
```

# Using a Birth-Death Process Tree
```{r}
Sim1a <- TraitSim(a = 0.98, b = 0.95, seed1 = 132, seed2 = 51,
                  tree.method = "bd")

png(filename="../figures/Simulation_bd3.png",
    width = 1600, height = 800, res = 96*2)

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(Sim1a$node.traits2$Offspring, cex = 0.25, frame = "circle", bg = Sim1a$n.col)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- Sim1a$edge[(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), ]
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), 
           pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(Sim1a$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim1a$t.col, adj = c(3, 0.5))
nodelabels(node = c(Sim1a$nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(Sim1a$nodes), text = c(Sim1a$nodes), cex = 0.5, frame = "none")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("../figures/Simulation_bd3.png")
grid.raster(img)
```

# Edge Length Distribution Around Traits
```{r}
Sim <- TraitSim(a = 0.965, b = 0.98, seed1 = 132, seed2 = 51,
                  tree.method = "bd")

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))

# Trait Evolution
plot(Sim$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(Sim$node.traits2$Offspring, cex = 0.25, frame = "circle", bg = Sim$n.col)
tiplabels(pch = 22, bg = Sim$t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- Sim$edge[(grep("*,[0-9][0-9][0-9]", rownames(Sim$edge))), ]
plot(Sim$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim$t.col, adj = c(3, 0.5))
edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim$edge))), 
           pie = edge2, cex = 0.75, piecol = c("gray", "red"))

# Trait Conservation
plot(Sim$tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = Sim$t.col, adj = c(3, 0.5))
nodelabels(node = c(Sim$nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(Sim$nodes), text = c(Sim$nodes), cex = 0.5, frame = "none")

```



```{r}

help("tree.bisse")


```

# Yule Tree
```{r}
# Parameters
birth = 0.2; a = 0.95; b = 0.98

# Generate Yule Tree
y.tree <-sim.bdtree(b = birth, d = 0, stop = "taxa", n = 100)
y.tree$tip.label <- paste("OTU", sprintf("%05d", seq(1:100)), sep = "")

# Save a few values
Ntips <- length(y.tree$tip.label)
Nedges <- y.tree$Nnode + Ntips

# Standardize Branch Lengths
tip.dist <- round(dist.nodes(y.tree)[1:Ntips, Ntips + 1], 5)
if (var(tip.dist) != 0){
  stop("Tree may not be ultrametric")
}
std.fac <- mean(tip.dist)
y.tree$edge.length <- (y.tree$edge.length / std.fac) * 100

# Generate Traits Matrix
traits <- matrix(NA, nrow = Nedges, ncol = 3)
colnames(traits) <- c("Parent", "Offspring", "Trait")
traits <- as.data.frame(traits)

# Define Trait States
traitNames <- c("Off","On")

# Define Root Ancestor Traits
traits[1,] <- c("None", "101", "Off")

# Run Trait Model Given the Tree
for (i in 1:(Nedges - 1)){
  traits[i + 1, 1:2] <- y.tree$edge[i, ]
  t <- y.tree$edge.length[i]
  p <- y.tree$edge[i,1]
  o <- y.tree$edge[i,2]
  index.p <- which(traits$Offspring == as.character(p))
  init <- traits$Trait[index.p]
  if (init == "Off"){u <- c(1,0)} else {
    if (init == "On"){u <- c(0,1)}}
  M <- MCtrait(t = t , a = a, b = b)
  prob <- round(u %*% M$M, 4)
  s <- sample(traitNames, size = 1, prob = prob)
  traits[i + 1, 1] <- p   # Parent
  traits[i + 1, 2] <- o   # Offspring
  traits[i + 1, 3] <- s   # Trait State
}

node.traits <- traits[which(as.numeric(traits$Offspring) > Ntips), ]
node.traits2 <- node.traits[order(as.numeric(node.traits$Offspring)), ]

tip.traits <- traits[which(as.numeric(traits$Offspring) <= Ntips), ]
tip.traits2 <- tip.traits[order(as.numeric(tip.traits$Offspring)), ]

n.col <- node.traits2$Trait
n.col <- gsub("On", "red", gsub("Off", "gray", n.col))

t.col <- tip.traits2$Trait
t.col <- gsub("On", "red", gsub("Off", "gray", t.col))

Obs.Traits <- data.frame(OTU = y.tree$tip.label, 
                                Traits = tip.traits$Trait)

# Trait Evolution
plot(y.tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(node.traits2$Offspring, cex = 0.25, frame = "circle", bg = n.col)
tiplabels(pch = 22, bg = t.col, adj = c(3, 0.5))
```

# Trait Description
```{r}
traits$P.trait <- NA
traits$P.trait[1] <- "Off"
for (i in 2:dim(traits)[1]){
  par.id <- which(traits$Offspring == traits$Parent[i])
  traits$P.trait[i] <- traits$Trait[par.id]
}
trait.evol <- traits[which(traits$P.trait == "Off" & traits$Trait == "On"), ]
trait.evol$distance <- dist.nodes(y.tree)[trait.evol$Offspring, Ntips + 1]

min.evol <- min(trait.evol$distance)



```

```{r}
# Trait Predictions with ASR

# Table of Observed Traits
x <- Obs.Traits$Traits; names(x) <- Obs.Traits$OTU

# Use ACE function for Baysian Posterior Probabilities
ASR <- ace(x, y.tree, type = "d", model = "ARD", CI = T,
           marginal = T, ip = c(1, 0))

# Use Simmap Function to run Markov Chain Monte Carlo for Trait Statse on Edges
# MCMC For Time in each state on each edge
ASR.2 <- make.simmap(y.tree, x, Q = "mcmc", model = "ARD", pi = c(1,0))
edge <- ASR.2$mapped.edge / tree$edge.length
pred <- c("Off", names(unlist(lapply(ASR.2$maps, tail, n = 1L))))


# Create Trait Prediction Data Frame
pred.traits <- data.frame(Parent = traits$Parent, Offspring = traits$Offspring)
pred.traits$Trait.Pred <- NA
pred.traits$Trait.Pred[1] <- "Off"

for (i in 2:dim(pred.traits)[1]){
  po.p <- strsplit(row.names(edge), ",")[[i - 1]]
  po.o <- as.vector(unlist(pred.traits[i + 1, 1:2]), mode = "character")
  if(all.equal(po.p, po.o) == FALSE){
    stop("predicted and observed trait dataframes do not match")
  }
  pred.traits$Trait.Pred[i] <- pred[i-1]
}

pred.node <- c("Off", names(unlist(lapply(ASR.2$maps, tail, n = 1L)))[1:98])
pred.node <- gsub("On", "red", gsub("Off", "gray", pred.node))
```


# Define Traits
x.2 <- as.numeric(gsub("Off", 0, gsub("On", 1, x)))
pred.traits <- data.frame(OTU = names(x), Traits1 = x.2, Traits2 = rev(x.2))                     
# Use ConsenTrait Function for Trait Conservation  
conserve <- ConsenTrait(tree, traits = traits, cutoff = 0.90)
nodes <- conserve$node[which(conserve$trait == 1)]



```{r}

layout(matrix(c(1:3), 1, 3))
par(mar = c(0.5, 0.5, 0.5, 0.5), oma = c(1, 1, 3, 1))


# Trait Evolution
plot(y.tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Simulation", side = 3, cex = 1, outer = F)
nodelabels(node.traits2$Offspring, cex = 0.25, frame = "circle", bg = n.col)
tiplabels(pch = 22, bg = t.col, adj = c(3, 0.5))

# ASR
# Limit Edges
edge2 <- edge[(grep("*,[0-9][0-9][0-9]", rownames(edge))), ]
plot(y.tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Ancestral State Reconstruction", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = t.col, adj = c(3, 0.5))
#edgelabels(edge = c(grep("*,[0-9][0-9][0-9]", rownames(Sim1a$edge))), 
#           pie = edge2, cex = 0.75, piecol = c("gray", "red"))
nodelabels(node.traits2$Offspring, cex = 0.25, frame = "circle", bg = pred.node)

# Trait Conservation
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 4, cex = 0.3)
mtext("Trait Conservation", side = 3, cex = 1, outer = F)
tiplabels(pch = 22, bg = t.col, adj = c(3, 0.5))
nodelabels(node = c(nodes), pch = 21, bg = "red", cex = 2.5)
nodelabels(node = c(nodes), text = c(nodes), cex = 0.5, frame = "none")

# Trait Evolution Off -> On


# Sim Traits Off -> On
trait.pred <- node.traits2
rownames(trait.pred) <- seq(1:dim(trait.pred)[1])
trait.pred$P.trait <- NA
trait.pred$P.trait[1] <- "Off"
for (i in 2:dim(trait.pred)[1]){
  par.id <- which(trait.pred$Offspring == trait.pred$Parent[i])
  trait.pred$P.trait[i] <- trait.pred$Trait[par.id]
}
pred.evols <- trait.pred[which(trait.pred$P.trait == "Off" & trait.pred$Trait == "On"), ]
pred.evols$distance <- dist.nodes(tree)[evols$Offspring, length(tree$tip.label) + 1]


```



help

