---
title: "Microbial Innovations: Simulations"
author: "Mario E Muscarella"
date: "Last updated on `r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
  - \usepackage{graphics}
output: 
  pdf_document: 
    fig_caption: yes
---

```{r setup, include=FALSE}
# Knitr Options
knitr::opts_chunk$set(echo = TRUE)

# Clear Environment and Set Working Directory
rm(list=ls())
setwd("~/GitHub/microbial-innovations/analyses")

# Load Packages
library("phytools")
library("ape")
library("mnormt")
library("MASS")
library("expm")
library("diagram")
library("adephylo")
```


# Mock Data: 
To test trait innovation approaches, we are going to work with some mock data. 
The goal is to have managable trees with traits that we can test various methods.

# Simulate Phylogenetic Tree of OTUs
```{r}
# Set Seed for Reproducibility
set.seed(149)

# Generate a tree
#tree <- pbtree(n = 100, scale = 1)
tree <- rcoal(n = 100, rooted = T)
tree$tip.label <- paste("OTU", sprintf("%05d", seq(1:100)), sep = "")

# Re Order Tree
tree <- reorder(tree)
```

# Siulate Traits Data using Markov Model for Trait Evolution
```{r}
a <- 0.7
b <- 0.4
Q <- matrix(c(a, 1 - a, b, 1 - b), 2, 2, byrow = T)

eig.Q <- eigen(Q)

# prob
twostep.trans <- Q %*% Q
init.dist = matrix(c(0.4, 0.6), 1, 2)
init.dist %*% twostep.trans


C <- eigen(Q)$vectors
D <- diag(eigen(Q)$values)

init.dist %*% (C %*% exp(D*0.5) %*% ginv(C))

```


```{r}
library(expm)
library(markovchain)
library(diagram)
library(pracma)

stateNames <- c("Rain","Nice","Snow")
Oz <- matrix(c(.5,.25,.25,.5,0,.5,.25,.25,.5),
             nrow=3, byrow=TRUE)
row.names(Oz) <- stateNames; colnames(Oz) <- stateNames
Oz


plotmat(Oz,pos = c(1,2),
        lwd = 1, box.lwd = 2,
        cex.txt = 0.8,
        box.size = 0.1,
        box.type = "circle",
        box.prop = 0.5,
        box.col = "light yellow",
        arr.length=.1,
        arr.width=.1,
        self.cex = .4,
        self.shifty = -.01,
        self.shiftx = .13,
        main = "")

Oz3 <- Oz %^% 3
round(Oz3,3)

u <- c(0, 1, 0)
round(u %*% Oz3,3)


```







# Trait Evolution
```{r}
# Create Tree

# Set Seed for Reproducibility
set.seed(199)

# Generate a tree
#tree <- pbtree(n = 100, scale = 1)
tree <- rcoal(n = 100, rooted = T)
tree$tip.label <- paste("OTU", sprintf("%05d", seq(1:100)), sep = "")

# Re Order Tree
tree <- reorder(tree)

plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 1, cex = 0.5)

# Generate Traits Under MC Model
set.seed(171)
str(tree)
summary(tree$edge.length)

tree$edge.length <- tree$edge.length * 100
summary(tree$edge.length)

int.trait <- rep(NA, tree$Nnode)
int.traits <- seq(from = 101, to = 100 + length(int.trait)) 

traits <- matrix(NA, nrow = (tree$Nnode + length(tree$tip.label)), ncol = 3)
colnames(traits) <- c("Parent", "Offspring", "Trait")
traits <- as.data.frame(traits)

traitNames <- c("Off","On")
traitStates <- c(0,1)
Q <- matrix(c(0.978, 0.012, 0.015, 0.985), 2, 2, byrow = T)

traits[1,] <- c("None", "101", "Off")

for (i in 1:dim(tree$edge)[1]){
  traits[i +1,1:2] <- tree$edge[i, ]
  t <- as.integer(round(tree$edge.length[i], 0))
  p <- tree$edge[i,1]
  o <- tree$edge[i,2]
  init <- traits$Trait[which(traits$Offspring == as.character(p))]
  if (init == "Off"){u <- c(1,0)} else {
    if (init == "On"){u <- c(0,1)}}
  prob <- u %*% (Q %^% t)
  s <- sample(traitNames, size = 1, prob = prob)
  traits[i + 1, 1] <- p 
  traits[i + 1, 2] <- o
  traits[i + 1, 3] <- s
}

node.traits <- traits[which(as.numeric(traits$Offspring) > length(tree$tip.label)), ]
node.traits2 <- node.traits[order(as.numeric(node.traits$Offspring)), ]

t.col <- node.traits2$Trait
t.col <- gsub("Off", "gray", t.col)
t.col <- gsub("On", "white", t.col)

plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 1, cex = 0.5)
nodelabels(int.traits, cex = 0.5, frame = "circle", bg = t.col)

```

# MCtraits Test Function
```{r}


a <- 0.7
b <- 0.55

t = 0.1

Q <- matrix(c(a, 1-a, 1-b, b), 2, 2, byrow = T)

x <- ((a - 1)/(a + b - 2)) * log(a + b - 1)
y <- ((b - 1)/(a + b - 2)) * log(a + b - 1)

H <- matrix(c(x, -x, -y, y), 2, 2, byrow = T)

C <- eigen(H)$vectors
D <- diag(eigen(H)$values)

C %*% ginv(C)
C %*% Matrix::solve(C)

Q2 <- Q %^% t
M <- C %*% expm(D * t) %*% ginv(C)
M2 <- expm(H * t)
rowSums(Q2)
rowSums(M)
rowSums(M2)

```

# Generate Tree
```{r}
# Set Seed for Reproducibility
set.seed(199)

# Generate a tree
#tree <- pbtree(n = 100, scale = 1)
tree <- rcoal(n = 100, rooted = T)
tree$tip.label <- paste("OTU", sprintf("%05d", seq(1:100)), sep = "")

# Re Order Tree
tree <- reorder(tree)
```

# Visulaize Tree
```{r}
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 1, cex = 0.5)
```

# Define Markov Chain Trait Model Function
```{r}
MCtrait <- function(t = 5, a = 0.7, b = 0.5){
  Q <- matrix(c(a, 1-a, 1-b, b), 2, 2, byrow = T)

  x <- ((a - 1)/(a + b - 2)) * log(a + b - 1)
  y <- ((b - 1)/(a + b - 2)) * log(a + b - 1)

  H <- matrix(c(x, -x, -y, y), 2, 2, byrow = T)
  C <- eigen(H)$vectors
  D <- diag(eigen(H)$values)

  M <- expm(H * t)
  M2 <- C %*% expm(D * t) %*% ginv(C)
  
  return(list(M = M2, Q = Q))
}
```

# Generate Traits Under Markov Chain Trait Model
```{r}
# Set Seed for Reproducibility
set.seed(171)

# Generate Traits Matrix
traits <- matrix(NA, nrow = (tree$Nnode + length(tree$tip.label)), ncol = 3)
colnames(traits) <- c("Parent", "Offspring", "Trait")
traits <- as.data.frame(traits)

# Define Trait States
traitNames <- c("Off","On")

# Define Root Ancestor Traits
traits[1,] <- c("None", "101", "Off")

# Run Trait Model Given the Tree
for (i in 1:dim(tree$edge)[1]){
  traits[i +1,1:2] <- tree$edge[i, ]
  t <- tree$edge.length[i]
  p <- tree$edge[i,1]
  o <- tree$edge[i,2]
  init <- traits$Trait[which(traits$Offspring == as.character(p))]
  if (init == "Off"){u <- c(1,0)} else {
    if (init == "On"){u <- c(0,1)}}
  M <- MCtrait(t = t , a = 0.5, b = 0.55)
  prob <- round(u %*% M$M, 4)
  s <- sample(traitNames, size = 1, prob = prob)
  traits[i + 1, 1] <- p 
  traits[i + 1, 2] <- o
  traits[i + 1, 3] <- s
}

node.traits <- traits[which(as.numeric(traits$Offspring) > length(tree$tip.label)), ]
node.traits2 <- node.traits[order(as.numeric(node.traits$Offspring)), ]

tip.traits <- traits[which(as.numeric(traits$Offspring) <= length(tree$tip.label)), ]
tip.traits2 <- tip.traits[order(as.numeric(tip.traits$Offspring)), ]

n.col <- node.traits2$Trait
n.col <- gsub("On", "white", gsub("Off", "gray", n.col))

t.col <- tip.traits2$Trait
t.col <- gsub("On", "white", gsub("Off", "gray", t.col))
```

# Plot MC Model Traits on Tree
```{r}
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 2, cex = 0.3)
nodelabels(node.traits2$Offspring, cex = 0.25, frame = "circle", bg = n.col)
tiplabels(pch = 22, bg = t.col, adj = c(1.5, 0.5))
```

# Conceptual Figure for MC Trait Model
```{r}
plotmat(M$Q,pos = c(2),
        lwd = 1, box.lwd = 2,
        cex.txt = 0.8,
        box.size = 0.1,
        box.type = "circle",
        box.prop = 0.5,
        box.col = "light yellow",
        arr.type = "triangle",
        arr.length= 0.3,
        arr.width= 0.2,
        self.cex = 0.4,
        self.shifty = 0.03,
        self.shiftx = c(-0.13, 0.13),
        shadow.size = 0.001,
        main = "",
        name = c("Off", "On"))
```


# Predict Traits with Ancestral State Reconstruction
```{r}
Obs.Traits <- data.frame(OTU = tree$tip.label, Traits = tip.traits2$Trait)

# Use ACE function for Baysian Posterior Probabilities
ASR <- ace(Obs.Traits$Traits, tree, type = "d", model = "ARD", CI = T,
           marginal = T)

# Use Simmap Function to run Markov Chain Monte Carlo for Trait Statse on Edges
# MCMC For Time in each state on each edge
x <- Obs.Traits$Traits; names(x) <- tree$tip.label
ASR.2 <- make.simmap(tree, x, Q = "mcmc", model = "ARD", pi = c(1,0))
edge <- ASR.2$mapped.edge / tree$edge.length
```

# Plot ASR Results
```{r}
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 2, cex = 0.3)

tiplabels(pch = 22, bg = t.col, adj = c(1.5, 0.5))

nodelabels(text = node.traits2$Offspring, cex = 0.5, frame = "none", adj = c(0.5,-2.2))
nodelabels(thermo = ASR$lik.anc, cex = 0.4, piecol = c("gray", "white"))

edgelabels(pie = edge, cex = 0.4, piecol = c("gray", "white"))

```

# Define Trait Conservation Function
```{r}
# Define Function
ConsenTrait <- function(tree = "", traits = "", cutoff = 0.9){

  # Import Tree and Root  if Needed
  if (is.rooted(tree) == FALSE){
    root_tree <- midpoint.root(tree)
  } else {
    root_tree <- tree
  }
  
  # Import Traits into Function
  table <- traits
  
  # Drop tips not found in trait table
  z <- subset(tree$tip.label, !(tree$tip.label %in% table[,1]) )
  if (length(z) > 0){
    drop.tip(tree,z)
  }
  
  # Replace any negative branch lengths
  root_tree$edge.length[root_tree$edge.length <= 0] = 0.00001
  
  # ID all subtrees
  subtree <- subtrees(root_tree, wait = FALSE)

  # Initializing Results Table
  y = rep(NA, (length(subtree) * (dim(table)[2] - 1)))
  cluster_size_tab <- data.frame(trait = NA, subtree = NA, node = NA, 
                                 distance = NA, cluster_size = NA)

  # Loop Through Traits
  for (i in 2:ncol(table)){
    print(paste("Analyzing Trait", i - 1, "of", 
                ncol(table)[[1]] - 1, "...", sep = " "), quote = F)
    
    # Make Temp Table
    table_tmp <- data.frame(ID = table[, 1], Trait = table[,i])

    # Remove All Entries Not in Tree and Sort by ID
    table2 <- table_tmp[which(table_tmp$ID %in% root_tree$tip.label), ]
    table2 <- table2[sort(table2$ID), ]
    
    # Initialize Temp Result Vectors
    positives <- vector(mode = "character", length = 0)
    cluster_size <- numeric(length=0)
    cluster_dist <- numeric(length = 0)
    node_positive <- vector(mode = "character", length = 0)
    
    # Loop through all subtrees and determine if any subtrees have >90% positives
    for (j in 1:length(subtree)){
      tip_names <- subtree[[j]]$tip.label
      if (mean(table2$Trait[which(table2$ID %in% tip_names)]) > cutoff){
        match_test <- match(tip_names, positives)
        if (all(is.na(match_test))){
          positives <- c(positives,tip_names)
          node_positive <- subtree[[j]]$node.label[1]
          cluster_dist <- distRoot(subtree[[j]], tip_names, method = c("p"))
          cluster_size <- c(cluster_size, mean(cluster_dist))

          cluster_size_tab[j + length(subtree) * (i - 2), ] <- c(i - 1, j, node_positive, 
                                     mean(cluster_dist), length(cluster_dist))
          
        } else {
          if (any(is.na(match_test))) {
            print("some NAs - something is weird")
          } 
        }
      }
    }
  }
  data.out <- cluster_size_tab[complete.cases(cluster_size_tab), ]
  return(data.out)
}
```

# Run Trait Conservation Function
```{r}
# Define Traits
x.2 <- as.numeric(gsub("Off", 0, gsub("On", 1, x)))
traits <- data.frame(OTU = names(x), Traits1 = x.2, Traits2 = rev(x.2))                        
conserve <- ConsenTrait(tree, traits = traits, cutoff = 0.75)
nodes <- conserve$node[which(conserve$trait == 1)]
```

# Plot Trait Conservation Results
```{r}
plot(tree, "c", FALSE, no.margin = TRUE, label.offset = 2, cex = 0.3)
tiplabels(pch = 22, bg = t.col, adj = c(1.5, 0.5))

# nodelabels(text = node.traits2$Offspring, cex = 0.5, frame = "none", adj = c(0.5,-2.2))

nodelabels(node = c(nodes), pch = 21, bg = "white", cex = 2.5)
nodelabels(node = c(nodes), text = c(nodes), cex = 0.5, frame = "none")
```







